---
title: "Golden Rules for Reproducible Statistical Analyses"
author: "Amy Gimma and Thibaut Jombart"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: pygments
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_collapse: no
    toc_depth: 2
    toc_float: yes
    css: !expr here::here('css', 'style.css')
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      fig.width = 8,
                      fig.height = 6,
                      dpi = 150,
                      warning = FALSE,
                      message = FALSE)
```

```{r read_scripts}

## read scripts
path_to_scripts <- here::here("scripts")

scripts_files <- dir(path_to_scripts, pattern = ".R$",
                     full.names = TRUE)
load_dictionary <- here::here("scripts", "load_dictionary.R")
for (file in scripts_files) if (!file == load_dictionary) source(file)

```



<!-- Rule 1 -->
# Organize your document logically {.tabset .tabset-fade .tabset-pills}

A clear structure is essential for readers to understand and navigate your
analysis document. We recommend the following structure:

* Data preparation
* Analysis of *xxx*
* Analysis of *yyy*
* ...
* Export outputs
* (Optionally) System information

Where the analyses of *xxx*, *yyy*, ... should have identical structures, if
possible.



## Data preparation

We recommend the following structure:

* **Load packages needed**

* **Load the raw data**

* **Clean the raw data** 
    + *standardise data* (e.g. `linelist::clean_data`)
    + *convert dates* that need converting (e.g. `linelist::guess_dates`)
    + *fix typos* (e.g. `linelist::clean_variable_spelling`)

* **Add new variables** (e.g. using `mutate`)

* **Subset entries** (rows) of the data (e.g. using `filter`)

* **Define custom colors** (e.g. using `scale_fill_manual` or `scale_color_manual`)



## Data analysis

Organise the work in a systematic way, using one sub-section per type of
analysis, and starting with general analyses before subsetted or stratified
ones.

The following workflow should apply to most cases:

* **Data manipulations and computations**: this part will derive numbers and metrics
  needed by the analysis
  
* **Graphics**: graphical display of the results; we recommend using *ggplot2* where
  possible

* **Table**: a table providing numbers matching the graphics, and useful summary
  statistics, e.g. means, medians, confidence intervals, etc.



## Export outputs

As a rule, we recommend to export every table displayed in the document to a
`.xlsx` and/or `.rds` files, so that they can be used in further analyses in R
(`.rds` file) or using other software. (`.xlsx` file). We recommend `.xlsx` over
text-based formats (e.g. `.txt`, `.csv`) as it preserves variable types.

The following code exports all tables named in `to_report` to `xslx` files,
stored inside the folder `produced_xlsx`; replace `xlsx` with `rds` for R files,
but in this case, do not insert the links as `rds` files cannot be opened within
a web browser (they are compressed binary representations of the data):

```{r xlsx_exports, eval = FALSE}

if (!dir.exists("produced_xlsx")) {
  dir.create("produced_xlsx")
}

to_export <- c("table_1",
               "table_xxx",
               "table_yyy",
               "table_zzz")

for (e in to_export) {
  rio::export(get(e),
              file.path("produced_xlsx",
                        paste0(e, ".xlsx")))
}

```

The following code will create links in your document pointing to the exported
files:

```{r links, results = "asis", eval = FALSE}

for (e in to_export) {
  txt <- sprintf("- [%s.xlsx](%s.xlsx)",
                 e,
                 file.path("produced_xlsx",
                           e))
  cat(txt, sep = "\n")
}

```


## System information

This is optional, but can be useful for audit purposes, or for diagnosing issues
in the results generated. We recommend including the following:

* `Sys.info()`: basic system information

* `R.version`: version of **R**

* `sessionInfo()`: which packages are loaded, and which versions are they?

* `params`: this list will contain optional parameters passed at compilation
  time through the `params` argument of `compile_report` or `update_reports`
  




<!-- Rule 4 -->
# Follow standard naming conventions {.tabset .tabset-fade .tabset-pills}

Predictable naming helps streamline writing and reviewing code
* *column names* should be lower snake case in utf-8 format without any special
characters or acccents (ex: Use;  `Unité âge (mois, années)` )




<!-- Rule 5 -->
# Show uncertainty {.tabset .tabset-fade .tabset-pills}

## Overview
Confidence intervals should be used: 

* to represet the uncertainty around generalisability from sample to population
* to represent uncertainty and variable conditions in data collection
* on predictions and imputations



```

 -------------------------    |Yes|----- **Include Confidence Interval**
|                         |     |  
|     Might the value     |     |
| observed be different   |-----|
| with a larger sample?   |     |                 |Yes|--| **CI not required**
|                         |     |    -----------   |
 -------------------------    |No|--| You sure? |--|
                                     -----------   |
                                                  |No|-- **Ask a statistician!**
```

## Example

Let us make some toy data, and plot a trend **without showing dispersion** in
the data:

```{r rule_5, fig.width = 4, fig.height = 4, fig.pos = "center"}

set.seed(1)
x <- rep(1:50, each = 15)
y <- rnorm(length(x), mean = 1)
df <- data.frame(x, y)

ggplot(df, aes(x = x, y = y)) +
  theme_bw() +
  geom_smooth() +
  labs(title = "Is there a trend here?")

```

Looking at this graph, **we may think there is a downward trend**, and start
interpreting what it means etc. **In fact, this is purely artifactual**. Data
have been generated in such a way that there is no correlation between the two
variables. This would be obvious if we add the data, thereby representing the
variability in the data:

```{r rule_5_2, fig.width = 4, fig.height = 4, fig.pos = "center"}

ggplot(df, aes(x = x, y = y)) +
  theme_bw() +
  geom_point() +
  geom_smooth() +
  labs(title = "No, there is not!")

```





<!-- Rule 6 -->
# Code systematically {.tabset .tabset-fade .tabset-pills}

## Overview

Be consistent in functions in libraries used

* *use `linelist::clean_data`* with a data dictionary for cleaning dates,
column names and values, etc. Take the time to understand how to use this 
package and how it works.

* for example choose between `n()`, `tally()`, and `count()` when summarizing
grouped data



## Example

Let's look at some basic code to count cases by location, gender and outcomes on
the MERS dataset from the `outbreaks` package.

```{r rule_6}

mers <- outbreaks::mers_korea_2015$linelist %>%
  as_tibble()

# 2x2 table 1: location / sex
mers %>%
  group_by(place_infect, sex) %>%
  summarise(n = n())

## 2x2 table 2: location / outcome
loc_out_tab <- count(mers, place_infect, outcome)
loc_out_tab

##################################
## 2x2 table 3: location / outcome
##################################
table_sex_outcome <- with(mers, table(sex, outcome)) %>% as.data.frame() %>% as_tibble
table_sex_outcome

```

The code is correct, but there are plenty of consistency issues here; to name a
few:

* using 3 different approaches to construct the 2x2 tables

* not using the piping operator ` %>% ` in table 2, but using it elsewhere

* annotations are not consistent

* not going back to a new line after the piping operators for table 3

* in table 3, not using parenthesis for `as_tibble`, but using them for
  `as.data.frame()`

* storing the 2nd and 3rd table as objects, but not for the 1st one

* using different naming conventions: table 2 is abbreviated and `tab` comes
  last, whilst table 3 is explicit and `table` comes first in the name
  
A more consistent version of the code would look like:

```{r rule_6_2}

## 2x2 table 1: location / sex
table_location_sex <- mers %>%
  count(place_infect, sex)
table_location_sex

## 2x2 table 2: location / outcome
table_location_outcome <- mers %>%
  count(place_infect, outcome)
table_location_outcome

## 2x2 table 3: location / outcome
table_sex_outcome <- mers %>%
  count(sex, outcome)
table_sex_outcome

```



<!-- Rule 7 -->
# Always use relative paths {.tabset .tabset-fade .tabset-pills}

Absolute paths will not work when the code is run on other computers. Use the
`here` library with R project files to establish a root directory

```{r eval = FALSE}

## NOT:
read.csv("P://user/specific/path/to/shared/directory/data/file_name.csv")

## Use:
read.csv(here::here("data", "file_name.csv")


```



<!-- Rule 8 -->
# Version and archive documents {.tabset .tabset-fade .tabset-pills}

Use semantic versioning in a clearly organized, standardized file structure

```
.
+-- linelist_investigations
|   +-- css
|   +-- data
|   +-- report_outputs
|   +-- report_sources
|       +-- aaa_clean_linelist_2019-29-09.Rmd
|       +-- epicurve_2019-04-10.Rmd
|       +-- _archive
|           +-- epicurve_2019-23-09.Rmd
|           +-- epicurve_2019-12-09.Rmd

```

<!-- Rule 9 -->
# Use descriptive naming {.tabset .tabset-fade .tabset-pills}

Variables, functions, and data sets should be named in a way that describes the 
object

<!-- Rule 10 -->
# Prepare presentation-ready plots {.tabset .tabset-fade .tabset-pills}

**Plots should be ready to copy and paste into an official presentation**

* check labels for correct spelling, accents, and grammar 
* uniform colors for variables throughout a document
* text size large
* avoid overlapping labels or other features


<!-- Rule 11 -->
# Always have your code reviewed {.tabset .tabset-fade .tabset-pills}

Ensure that:

* Analysis is appropriate to setting and underlying data is well understood, 
local collaborators should be included and acknowledged
* statistical methodology is applied correctly
* R code is readable and correctly implements methodology 
* markdown and the report document are well organized and properly formatted

<!-- Rule 12 -->
# Keep data safe {.tabset .tabset-fade .tabset-pills}

* Respect the privacy of the individuals described in the data

* Never label data as anonomysed without complying with best practices of the
standards and methods of anonymyzation (and clear final product with a 
qualified statistician)
